A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-10 paper/paper.tex
more edits to set connectives section

@@

Next, in Figure~\ref{fig:set-connectives}, we cover the connectives for
the sets/profunctors.  In these constructions, the sets/profunctors act
like types, and they are given introduction and elimination rules for
elements/natural transformations.
%
The top of the figure includes the formation rules for tensor products
$\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
\alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$, and
the cartesian unit and product types $1$ and $R \times S$.  
%
The remainder of the figure gives the introduction and elimination rules
for these sets (the $\beta/eta$-rules are in the appendix).  
%
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$.
%
Its introduction and elimination rules are like those of an
\emph{ordered} variant of equality: the introduction rule is the
identity morphism (reflexivity) and the elimination rule is an
induction principle that is a variant of the J rule of equality
elimination: we can use a term of $s : \punitinXfromYtoZ {\cat C} a b$
by specifying the behavior when $s$ is of the form $\punitrefl \alpha$
in the form of a continuation $\alpha. t$. Next, the tensor product of
sets is a kind of combined existential quantifier and monoidal
product, which we combine into a single notation
$\tensorexistsXwithYandZ \beta P Q$, where $\beta$ is the covariant
variable of $P$ and the contravariant variable of $Q$. Then the
covariant variable of the tensor product is the covariant variable of
$Q$ and the contravariant variable similarly comes from $P$.
%
In ordinary category theory, this is the \emph{composition} of
profunctors, and is defined by a coend of a product. We require that
the variable $\beta$ quantifies over a small category $\cat D$, as in
general this composite doesn't exist for large categories.
%
The introduction and elimination are like those for a combined tensor
product and existential type: the introduction rule is a pair of
terms, with an appropriate instantiation of $\beta$, and the
elimination rule says to use a term of a tensor product, it is
sufficient to specify the behavior on two elements typed with an
arbitrary middle object $\beta$.

FIXME: Note that in the unit elimination, the continuation is typed with
a context consisting of only a single object $\alpha$, in contrast with
normal rules for a unit type in natural deduction for ordered linear
logic such as \cite{polakawpfenning}, which would split the context
$\Phi$ into $\Phi_l \jnctx \Phi_m \jnctx \Phi_r$ where $\Phi_m$ is used
by the scrutinee and $\Phi_l \jnctx \Phi_r$ is used by the
continuation. However, unlike in ordered linear logic, the $\Phi_l
\jnctx \Phi_r$ is not necessarily well formed because there is no reason
in general that $\Phi_m$ has contravariant and covariant dependence on
the same category.
%
This is analogous to a similar situation in dependent type theory for
identity types: TODO: Dan please help with this explanation!!

@@

Finally, we have the cartesian unit and product sets, which are
analogous to the normal unit and product of types. The most notable
point to emphasize is that in the formation rule for the product, the
two subformulae should have the same covariant and contravariant
dependence. So as with linear logic, the cartesian unit and product
complicate the syntactic formulation of linearity: the unit
constructor ``uses'' all variables and the product ``uses'' all
variables in both cases.

@@

    \text{Unit/morphism set:}\\
    \inferrule
  {\Gamma \pipe d^-\Xi \vdash a_1 : \cat C\\\\ \Gamma \pipe d^+ \Xi \vdash a_2 : \cat C}
  {\Gamma \pipe \Xi \vdash \punitinXfromYtoZ {\cat C} {a_1} {a_2}}
  \quad
    \inferrule
    {\Gamma\pipe\beta: \cat D \vdash a : \cat C}
    {\Gamma \pipe\cdot \vdash \punitrefl a : \punitinXfromYtoZ {\cat C} a a}
    \quad
%%     \inferrule*[right=Unit Elim]
%%     {\Gamma\vdash Q : \harr {\cat C} {\cat C}\and
%%      \Gamma\vdash t : \pendallXdotY {\alpha:\cat C} {\harrapp Q \alpha \alpha}}
%%     {\Gamma \vdash \punitelimtoYwithkontZ Q N :
%%       \pendallXdotY {\alpha_1} {\homrallXYtoZ {\alpha_2} {\punitinXfromYtoZ {\cat C} {\alpha_1}{\alpha_2}} {\harrapp Q {\alpha_1}{\alpha_2}}}}
    \inferrule
    {\Gamma\pipe \alpha:\cat C\vdash t : P[\alpha/\alpha^-;\alpha/\alpha^+]\\\\
     \Gamma\pipe\Phi \vdash s : \punitinXfromYtoZ {\cat C} A B}
    {\Gamma \vdash \punitelimkontZatABC {\alpha. t} A s B : P[A/\alpha^-;B/\alpha^+]}
    \\
  %% \inferrule*[right=Tensor intro function]
  %% {}
  %% {\alpha : \cat C, q : Q, \beta : \cat D, r : R, \gamma : \cat E \vdash (q, \beta,r) : \exists \beta : \cat D. Q \odot R}

@@

    \\
    \text{Tensor product:}\\
    \inferrule
    {\cat D \isSmall\\\\
      \Gamma \pipe d^-\Xi; \beta:D\vdash P \isaSet\\\\
     \Gamma \pipe \beta:D;d^+\Xi \vdash Q \isaSet}
    {\Gamma \pipe \Xi \vdash \tensorexistsXwithYandZ {\beta:D} P Q}
    \quad
  \inferrule
  {\Gamma \pipe d^+\Psi_s \vdash b : \cat D\\\\
   \Gamma \pipe \Psi_s \vdash s : P[b/\beta] \\\\
   \Gamma \pipe \Psi_t \vdash t : Q[b/\beta]}
  {\Gamma\pipe\Psi_s \jnctx \Psi_t\vdash \tensorintroatXwithYandZ b s t : \tensorexistsXwithYandZ {\beta : \cat D} P Q}
  \quad
  \inferrule
  {
   \Gamma\pipe \Phi_m \jnctx x:P, \beta:\cat D, y:Q \jnctx \Phi_r \vdash t : R\\\\
   \Gamma \pipe\Phi_m \vdash s : \tensorexistsXwithYandZ {\beta} P Q}
  {\Gamma\pipe\Phi_l\jnctx \Phi_m \jnctx \Phi_r \vdash \tensorelimWkontZ s {x.\beta.y. t} : R}
  \\
  %% \tensorintroatXwithYandZ B M N [\phi,B'/\beta',\psi] = \tensorintroatXwithYandZ {B[B'\beta]} {M[\phi]} {N[\psi]}

@@

  \\
  \text{Right hom:} \quad
  \inferrule
  {d^+\Xi \isSmall\\\\ \Gamma \pipe d^+\Xi; \alpha : \cat C \vdash  R \\\\ \Gamma \pipe  d^-\Xi; \alpha : \cat C \vdash P}
  {\Gamma \pipe \Xi \vdash \homrallXYtoZ {\alpha:\cat C} R P}
  \quad
  \inferrule
  {\Gamma \pipe \Phi, x : R, \alpha : \cat C \vdash t : P}
  {\Gamma \pipe \Phi \vdash
   \homrlambdaXatYdotZ {x : R} {\alpha : \cat C} t : \homrallXYtoZ \alpha R P}
  \quad
  \inferrule
  {\Gamma\pipe\Phi_f \vdash s : \homrallXYtoZ {\alpha:\cat C} R P \\\\
    d^+\Phi_a \vdash a : \cat C\\\\
    \Phi_a \vdash t : R[a/\alpha]
  }
  {\Gamma\pipe \Phi_f \jnctx \Phi_a \vdash \homrappXtoYatZ s t a : P[a/\alpha]}\\
  \quad

@@

  \\ 
  \text{Left hom:} \quad
  \inferrule
    {d^-\Xi \isSmall\\\\
      \Gamma \pipe \alpha:\cat C; d^-\Xi\vdash R \\\\ \Gamma \pipe \alpha : \cat C; d^+\Xi\vdash P}
    {\Gamma \pipe \Xi \vdash \homlallXYtoZ {\alpha: \cat C} R P}
  \quad
  \inferrule
  {\Gamma\pipe\alpha:\cat C, x : R, \Phi \vdash t : P}
  {\Gamma\pipe\Phi \vdash \homllambdaXatYdotZ {x : R} {\alpha : \cat C} t : \homlallXYtoZ \alpha R P}
  \quad
  \inferrule
  {\Gamma \pipe d^-\Phi_a \vdash a : \cat C\\\\
   \Gamma \pipe \Phi_a \vdash s : R[a/\alpha] \\\\
   \Gamma \pipe \Phi_f \vdash t : \homlallXYtoZ {\alpha:\cat C} R P}
  {\Gamma \pipe \Phi_a \jnctx \Phi_f \vdash \homlappXtoYatZ t s a : P[a/\alpha]}\\
  % tensor,
  % hom,
  % cartesian products,
  %

@@

    FIXME missing intro/elim rules for smallcat and cat

@@

>>>>

\newcommand{\quoth}[1]{\lceil{} {#1}\rceil{}}
\newcommand{\unquoth}[1]{\lfloor{} {#1}\rfloor{}}

@@

Next, in Figure~\ref{fig:set-connectives}, we cover the connectives for
the sets/profunctors, which classify elements/natural transformations
(the $\beta/eta$-rules are in the appendix).
%% %
%% The top of the figure includes the formation rules for tensor products
%% $\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
%% \alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
%% the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$, and
%% the cartesian unit and product types $1$ and $R \times S$.  
%% %
%% The remainder of the figure gives the introduction and elimination rules
%% for these sets .  
%% %
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$, which is well-formed when given two objects of the
same category.  
%
Its introduction and elimination rules are analogous to the usual rules
for equality in Martin-L\"of type theory.  The introduction rule is the
identity morphism (reflexivity) and the elimination rule is an induction
principle: we can use a term of $s : \punitinXfromYtoZ {\cat C} A B$ by
specifying the behavior when $s$ is of the form $\punitrefl \alpha$ in
the form of a continuation $\alpha. t$.  Like the $J$ elimination rule
for equality in Martin-L\"of type theory, $P$ must be ``fully general'',
i.e. well-typed for variables $\alpha$ and $\beta$.  Those familiar with
linear logic~\cite{polakawpfenning} might expect a more general rule,
with the continuation $t$ typed in a context $\Phi_l \jnctx \Phi_r$ and
$\Phi_l \jnctx \Phi \jnctx \Phi_r$ as the context of the conclusion of
the rule.  Because of dependency, this is not necessarily well-formed in
cases where the endpoints $A$ and $B$ of $\punitinXfromYtoZ{\cat C} A B$
are not fully general (i.e. not distinct variables).  However, the
instances of this more general rule that do type check are derivable
from our more restricted rule using right/left-hom types.

The tensor product of sets is a kind of combined existential quantifier
and monoidal product, which we combine into a single notation
$\tensorexistsXwithYandZ \beta P Q$, where $\beta$ is the covariant
variable of $P$ and the contravariant variable of $Q$. Then the
covariant variable of the tensor product is the covariant variable of
$Q$ and the contravariant variable similarly comes from $P$.
%
In ordinary category theory, this is the \emph{composition} of
profunctors, and is defined by a coend of a product. We require that
the variable $\beta$ quantifies over a small category $\cat D$, as in
general this composite doesn't exist for large categories.
%
The introduction and elimination are like those for a combined tensor
product and existential type: the introduction rule is a pair of
terms, with an appropriate instantiation of $\beta$, and the
elimination rule says to use a term of a tensor product, it is
sufficient to specify the behavior on two elements typed with an
arbitrary middle object $\beta$.

@@

Finally, we have the cartesian unit and product sets, which are
analogous to the normal unit and product of types. The most notable
point to emphasize is that in the formation rule for the product, the
two subformulae should have the same covariant and contravariant
dependence. So, as with linear logic, some constructions can
syntactically use a variable more than once and still be ``linear'': the
unit constructor ``uses'' all variables and the product ``uses'' all
variables in both cases.

@@

    \text{Unit/morphism set:}\\
    \inferrule
  {\Gamma \pipe d^-\Xi \vdash a_1 : \cat C\\\\ \Gamma \pipe d^+ \Xi \vdash a_2 : \cat C}
  {\Gamma \pipe \Xi \vdash \punitinXfromYtoZ {\cat C} {a_1} {a_2} \isaSet}
  \quad
    \inferrule
    {\Gamma\pipe\beta: \cat D \vdash a : \cat C}
    {\Gamma \pipe \beta : \cat D \vdash \punitrefl a : \punitinXfromYtoZ {\cat C} a a}
    \quad
%%     \inferrule*[right=Unit Elim]
%%     {\Gamma\vdash Q : \harr {\cat C} {\cat C}\and
%%      \Gamma\vdash t : \pendallXdotY {\alpha:\cat C} {\harrapp Q \alpha \alpha}}
%%     {\Gamma \vdash \punitelimtoYwithkontZ Q N :
%%       \pendallXdotY {\alpha_1} {\homrallXYtoZ {\alpha_2} {\punitinXfromYtoZ {\cat C} {\alpha_1}{\alpha_2}} {\harrapp Q {\alpha_1}{\alpha_2}}}}
    \inferrule
    {\Gamma\pipe \alpha:\cat C; \beta:\cat C \vdash P \isaSet \\\\
     \Gamma\pipe \alpha:\cat C\vdash t : P[\alpha/\alpha;\alpha/\beta]\\\\
     \Gamma\pipe\Phi \vdash s : \punitinXfromYtoZ {\cat C} A B}
    {\Gamma \pipe \Phi \vdash \punitelimkontZatABC {\alpha. t} A s B : P[A/\alpha;B/\beta]}
    \\
  %% \inferrule*[right=Tensor intro function]
  %% {}
  %% {\alpha : \cat C, q : Q, \beta : \cat D, r : R, \gamma : \cat E \vdash (q, \beta,r) : \exists \beta : \cat D. Q \odot R}

@@

    \\
    \text{Tensor product:}\\
    \inferrule
    {\cat D \isSmall\\\\
      \Gamma \pipe d^-\Xi; \beta:D\vdash P \isaSet\\\\
     \Gamma \pipe \beta:D;d^+\Xi \vdash Q \isaSet}
    {\Gamma \pipe \Xi \vdash \tensorexistsXwithYandZ {\beta:D} P Q \isaSet}
    \quad
  \inferrule
  {\Gamma \pipe d^+\Psi_s \vdash b : \cat D\\\\
   \Gamma \pipe \Psi_s \vdash s : P[b/\beta] \\\\
   \Gamma \pipe \Psi_t \vdash t : Q[b/\beta]}
  {\Gamma\pipe\Psi_s \jnctx \Psi_t\vdash \tensorintroatXwithYandZ b s t : \tensorexistsXwithYandZ {\beta : \cat D} P Q}
  \quad
  \inferrule
  {
   \Gamma\pipe \Phi_m \jnctx x:P, \beta:\cat D, y:Q \jnctx \Phi_r \vdash t : R\\\\
   \Gamma \pipe\Phi_m \vdash s : \tensorexistsXwithYandZ {\beta} P Q}
  {\Gamma\pipe\Phi_l\jnctx \Phi_m \jnctx \Phi_r \vdash \tensorelimWkontZ s {x.\beta.y. t} : R}
  \\
  %% \tensorintroatXwithYandZ B M N [\phi,B'/\beta',\psi] = \tensorintroatXwithYandZ {B[B'\beta]} {M[\phi]} {N[\psi]}

@@

  \\
  \text{Right hom:} \quad
  \inferrule
  {d^+\Xi \isSmall\\\\ \Gamma \pipe d^+\Xi; \alpha : \cat C \vdash  R
    \isaSet \\\\ \Gamma \pipe  d^-\Xi; \alpha : \cat C \vdash P \isaSet}
  {\Gamma \pipe \Xi \vdash \homrallXYtoZ {\alpha:\cat C} R P \isaSet}
  \quad
  \inferrule
  {\Gamma \pipe \Phi, x : R, \alpha : \cat C \vdash t : P}
  {\Gamma \pipe \Phi \vdash
   \homrlambdaXatYdotZ {x : R} {\alpha : \cat C} t : \homrallXYtoZ \alpha R P}
  \quad
  \inferrule
  {\Gamma\pipe\Phi_f \vdash s : \homrallXYtoZ {\alpha:\cat C} R P \\\\
    d^+\Phi_a \vdash a : \cat C\\\\
    \Phi_a \vdash t : R[a/\alpha]
  }
  {\Gamma\pipe \Phi_f \jnctx \Phi_a \vdash \homrappXtoYatZ s t a : P[a/\alpha]}\\
  \quad

@@

  \\ 
  \text{Left hom:} \quad
  \inferrule
    {d^-\Xi \isSmall\\\\
      \Gamma \pipe \alpha:\cat C; d^-\Xi\vdash R \isaSet \\\\ \Gamma
      \pipe \alpha : \cat C; d^+\Xi\vdash P \isaSet }
    {\Gamma \pipe \Xi \vdash \homlallXYtoZ {\alpha: \cat C} R P \isaSet}
  \quad
  \inferrule
  {\Gamma\pipe\alpha:\cat C, x : R, \Phi \vdash t : P}
  {\Gamma\pipe\Phi \vdash \homllambdaXatYdotZ {x : R} {\alpha : \cat C} t : \homlallXYtoZ \alpha R P}
  \quad
  \inferrule
  {\Gamma \pipe d^-\Phi_a \vdash a : \cat C\\\\
   \Gamma \pipe \Phi_a \vdash s : R[a/\alpha] \\\\
   \Gamma \pipe \Phi_f \vdash t : \homlallXYtoZ {\alpha:\cat C} R P}
  {\Gamma \pipe \Phi_a \jnctx \Phi_f \vdash \homlappXtoYatZ t s a : P[a/\alpha]}\\
  % tensor,
  % hom,
  % cartesian products,
  %

@@

\begin{figure}
  \begin{mathpar}
    \inferrule
    {~}
    {\Gamma\vdash \smallCats}

    \inferrule
    {~}
    {\Gamma\vdash \Cats}

@@

    \inferrule
    {\Gamma \vdash \cat C \isSmall}
    {\Gamma \vdash \quoth {\cat C} : \smallCats}

    \inferrule
    {\Gamma \vdash M : \smallCats}
    {\Gamma \vdash \unquoth M \isSmall}

    \inferrule
    {\Gamma \vdash \cat C \isaCat}
    {\Gamma \vdash \quoth {\cat C} : \Cat}

    \inferrule
    {\Gamma \vdash M : \Cat}
    {\Gamma \vdash \unquoth M \isaCat}
